Reasoning About a Machine with Local Capabilities

### 2019-02-10 toplas/paper.tex
[TOPLAS] Update malloc specification secion

\begin{toplas}
  \section{Malloc}
  \label{sec:malloc}
  In the examples we present in \sectionname~\ref{sec:examples}, we will assume
  the existence of a trusted malloc routine so that both the trusted code and
  the adversary can allocate new memory. Malloc is considered part of the
  trusted computing base, as mentioned in
  \sectionname~\ref{sec:capab-mach-with}. This is unavoidable: if we cannot
  trust malloc, then we cannot use the memory it allocates as we have no idea
  whether it is aliased by some untrusted program.

  Our semantic model is not specific to a particular implementation of malloc,
  so rather than providing the implementation, we provide a specification that
  malloc must satisfy. The specification expresses standard expectations of how
  malloc should behave, but making it realistic requires using some of the
  technical machinery from the logical relation. As such, this section is a bit
  technical and can safely be skipped on first read. The malloc specification is
  presented in \figurename~\ref{fig:malloc-spec}, and in the following, we will
  connect the description of the specification by referring to the item
  described.

We require a global capability for
invoking malloc (because if the capability were local, then a program with
access to malloc would have to give up this access when invoking untrusted
code, \ref{item:malloc-spec:eg}). The capability is assumed to have enter
permission (\ref{item:malloc-spec:eg}), so that malloc can
protect its internal state even when the capability is shared with untrusted
code.

In addition to these syntactic requirements, we also specify standard
expectations for how malloc behaves. Intuitively, we require that when malloc is
invoked with a length argument, then it will return a capability for a fresh
piece of memory of that size. It should be fresh in the sense that malloc has
not already given out a capability for any part of that memory before and will
not do so in the future. Also, when invoked with a nonsensical length argument
such as a negative integer or a capability, malloc should simply fail. However,
formulating these requirements is harder than one might expect. The problem is
that a realistic implementation of malloc needs to rely on internal state that
changes after every invocation and relies on invariants on that state. We can
only expect that malloc behaves according to its specification, if its internal
state satisfies the current state of the invariants in an executing system. To
express this, we will use the semantic model defined in
Section~\ref{sec:logical-relation}.

To allow malloc implementations to rely on internal state and invariants for
that state, we assume an initial region for malloc (\ref{item:malloc-spec:reg}). The region is assumed to be
permanent (since safety of the global malloc capability will depend on its
presence, \ref{item:malloc-spec:perm}).
% DOMI: I don't understand what the following uncommented text means
% This also means that the malloc specification should guarantee that
% use of malloc is safe at any point of execution no matter who has invoked it. To
% this end, the requirement on the initial malloc region should also hold true in
% any private future world.
% priv future means that malloc should not depend on local capabilities?
Furthermore, we want to express that malloc does not depend on any other memory
than its own internal state. This is expressed by a restriction on the malloc
region's state interpretation function, which (as explained in
\sectionname~\ref{subsec:worlds}) defines what memory segments it permits in a
given world. We require that the accepted memory segments only depend on the
presence of the malloc region itself, i.e.\ that in any world, the same memory
segments are accepted if we remove all regions except the malloc region. This
property should continue to hold throughout execution, so it must hold true for
any private future region of the initial malloc region (\ref{item:malloc-spec:independent}).

The malloc specification then dictates what malloc should do when invoked in a
memory with its internal state valid according to the malloc region (some future
evolution of the initial malloc region). If malloc is invoked with an invalid
length argument (that is, a negative integer or a capability), then we simply
require malloc to fail (\ref{item:malloc-spec:fail}). This part of the
specification does not actually rely on the malloc region: for simplicity, we
assume malloc does not need its internal state to check the argument. When
malloc is invoked with a valid length (\ref{item:malloc-spec:correct-exec}),
then it should return a fresh memory segment of the correct length. This segment
is required to come from the footprint of malloc, i.e.\ the memory owned by the
malloc region before the call. After malloc returns, we require the malloc
region to have evolved (according to the public future world relation) to a new
state where the new footprint is disjoint from the allocated memory. This
implies that future invocations of malloc can never return previously-allocated
memory.

@@

\begin{figure}[htb]
  \centering
  \begin{definition}[Malloc Specification]
    \label{spec:malloc}
    $c_\malloc$ satisfies the specification for malloc iff the following
    conditions hold:
    \begin{enumerate}
    \item \label{item:malloc-spec:eg} $c_\malloc = ((\entry,\glob),\_,\_,\_)$
    \item \label{item:malloc-spec:reg} There exists a $\iota_{\malloc,0}$ such that
      \begin{enumerate}
      \item \label{item:malloc-spec:perm} $\iota_{\malloc,0}.v = \perma$
      \item \label{item:malloc-spec:independent} For all $\iota' \futurestr \iota_{\malloc,0}, W,i$ with
        $W(i)=\iota'$, we have that
        \begin{equation*}
          \iota'.H (\iota'.s) (\xi^{-1}(W)) = \iota'.H (\iota'.s) (\xi^{-1}([i \mapsto W(i)]))
        \end{equation*}
      \item \label{item:malloc-spec:correct-exec} For all $\Phi \in \ExecConfs$, $\ms_{\var{footprint}}, \hsframe \in
        \HeapSegments$, $i, n, \size \in \nats$, $w_{\var{ret}} \in \Words$,
        $\iota_\malloc \futurestr \iota_{\malloc,0}$, we have that
        \[
          \begin{array}{l}
            \text{If }
            \memheap = \ms_{\var{footprint}} \uplus \hsframe \land \heapSat[\ms_{\var{footprint}}]{n}{[i \mapsto \iota_\malloc]} \land{} \\
            \quad\memreg(r_1) = \size \land \size \geq 0 \land  \memreg(r_0) = w_{\var{ret}} \land{} \\
            \quad \memreg(\pcreg) = \updatePcPerm{c_\malloc} \\ %and this needs to change with ABI.
            \text{Then, } \\
            \quad \exists \Phi' \in \ExecConfs \ldotp \exists \ms_{\var{footprint}}', \ms_{\var{alloc}} \in \HeapSegments\ldotp\\
            \qquad \exists j \in \nats \ldotp j > 0 \land \exists b',e'\in \Addrs \ldotp \exists \iota_\malloc' \in \Regions \ldotp \\
            \qquad \quad \Phi \step[j] \Phi' \land{} \\
            \qquad \quad \memheap[\Phi']=\ms_{\var{footprint}}' \uplus \hs_{\var{alloc}} \uplus \hsframe \land{}\\
            \qquad \quad \iota_{\malloc}' \futurewk \iota_\malloc \land{} \\
            \qquad \quad \heapSat[\ms_{\var{footprint}}']{n-j}{[i \mapsto \iota_\malloc']} \land{} \\
            \qquad \quad \dom(\hs_{\var{alloc}}) = [b',e'] \land \forall a \in [b',e']\ldotp \hs_{\var{alloc}}(a) = 0  \land{} \\
            \qquad \quad \memreg[\Phi'] = \memreg[\Phi]\update{\pcreg}{\updatePcPerm{w_{\var{ret}}}}\update{r_1}{((\rwx,\glob),b',e',b')} \land{} \\
            \qquad \quad \size - 1 = e'-b'
          \end{array}
        \]
      \item \label{item:malloc-spec:fail} For all $\Phi \in \ExecConfs$,
        \[
          \begin{array}{l}
            \text{If } (\memreg(r_1) \notin \ints \vee \memreg(r_1) < 0) \land
            \memreg(\pcreg) = \updatePcPerm{c_\malloc} \\
            \text{Then } \exists j \in \nats \ldotp \Phi \step[j] \failed
          \end{array}
        \]
      \end{enumerate}
    \end{enumerate}
  \end{definition}

  \caption{The specification of malloc.}
  \label{fig:malloc-spec}
\end{figure}
As described previously, the specification of malloc ensures that malloc has no
capabilities pointing \emph{out} of malloc. It does not, however, say anything about
capabilities that point \emph{in} to malloc. We obviously cannot
allow this if we want to trust malloc. We have chosen to keep the malloc
specification simple and let this assumption be in the lemmas that use malloc.
It is sufficient for these lemmas to require that there are no outside
capabilities for malloc in the initial configuration as capabilities cannot
appear out of thin air and the malloc specification makes sure that capabilities
are not leaked.

Malloc should not just be available to trusted programs, but also to possibly
malicious programs. This is safe as it follows from the specification that the malloc
capability is always safe in a world with the malloc region:
\begin{lemma}[Malloc is safe to pass to adversary]
  \label{lem:malloc-in-vr}
  For all $c_\malloc$ that satisfies the specification for malloc with region $\iota_{\malloc,0}$,  if $W(r) \futurestr \iota_{\malloc,0}$, then
  $\npair{c_\malloc} \in \stdvr(W)$ for all $n$.
\end{lemma}

>>>>

\begin{toplas}
  \section{Malloc}
  \label{sec:malloc}
  In the examples presented in \sectionname~\ref{sec:examples}, we will assume the existence of a trusted malloc routine, so that both the trusted code and the adversary are able to allocate new memory.
  Malloc is considered part of the trusted computing base as mentioned in \sectionname~\ref{sec:capab-mach-with}.
  This is unavoidable: if we cannot trust malloc, then we cannot use the memory it allocates as we have no idea whether it is aliased by some untrusted program.

  Our semantic model is not specific to a particular implementation of malloc, so rather than providing the implementation we provide a malloc specification.
  The specification expresses standard expectations about the behaviour of malloc, but making the specification realistic requires some of the technical machinery from the logical relation.
  As such, this section is a bit technical and can safely be skipped on first read.
  Definition~\ref{spec:malloc} is the malloc specification.
  Following the definition, we provide an informal description of the definition with references to each of the items in the definition.

\begin{definition}[Malloc Specification]
  \label{spec:malloc}
  $c_\malloc$ satisfies the specification for malloc iff the following
  conditions hold:
  \begin{enumerate}
  \item \label{item:malloc-spec:eg} $c_\malloc = ((\entry,\glob),\_,\_,\_)$
  \item \label{item:malloc-spec:reg} There exists a $\iota_{\malloc,0}$ such that
    \begin{enumerate}
    \item \label{item:malloc-spec:perm} $\iota_{\malloc,0}.v = \perma$
    \item \label{item:malloc-spec:independent} For all $\iota' \futurestr \iota_{\malloc,0}, W,i$ with
      $W(i)=\iota'$, we have that
      \begin{equation*}
        \iota'.H (\iota'.s) (\xi^{-1}(W)) = \iota'.H (\iota'.s) (\xi^{-1}([i \mapsto W(i)]))
      \end{equation*}
    \item \label{item:malloc-spec:correct-exec} For all $\Phi \in \ExecConfs$, $\ms_{\var{footprint}}, \hsframe \in
      \HeapSegments$, $i, n, \size \in \nats$, $w_{\var{ret}} \in \Words$,
      $\iota_\malloc \futurestr \iota_{\malloc,0}$, we have that
      \[
        \begin{array}{l}
          \text{If }
          \memheap = \ms_{\var{footprint}} \uplus \hsframe \land \heapSat[\ms_{\var{footprint}}]{n}{[i \mapsto \iota_\malloc]} \land{} \\
          \quad\memreg(r_1) = \size \land \size \geq 0 \land  \memreg(r_0) = w_{\var{ret}} \land{} \\
          \quad \memreg(\pcreg) = \updatePcPerm{c_\malloc} \\ %and this needs to change with ABI.
          \text{Then, } \\
          \quad \exists \Phi' \in \ExecConfs \ldotp \exists \ms_{\var{footprint}}', \ms_{\var{alloc}} \in \HeapSegments\ldotp\\
          \qquad \exists j \in \nats \ldotp j > 0 \land \exists b',e'\in \Addrs \ldotp \exists \iota_\malloc' \in \Regions \ldotp \\
          \qquad \quad \Phi \step[j] \Phi' \land{} \\
          \qquad \quad \memheap[\Phi']=\ms_{\var{footprint}}' \uplus \hs_{\var{alloc}} \uplus \hsframe \land{}\\
          \qquad \quad \iota_{\malloc}' \futurewk \iota_\malloc \land{} \\
          \qquad \quad \heapSat[\ms_{\var{footprint}}']{n-j}{[i \mapsto \iota_\malloc']} \land{} \\
          \qquad \quad \dom(\hs_{\var{alloc}}) = [b',e'] \land \forall a \in [b',e']\ldotp \hs_{\var{alloc}}(a) = 0  \land{} \\
          \qquad \quad \memreg[\Phi'] = \memreg[\Phi]\update{\pcreg}{\updatePcPerm{w_{\var{ret}}}}\update{r_1}{((\rwx,\glob),b',e',b')} \land{} \\
          \qquad \quad \size - 1 = e'-b'
        \end{array}
      \]
    \item \label{item:malloc-spec:fail} For all $\Phi \in \ExecConfs$,
      \[
        \begin{array}{l}
          \text{If } (\memreg(r_1) \notin \ints \vee \memreg(r_1) < 0) \land
          \memreg(\pcreg) = \updatePcPerm{c_\malloc} \\
          \text{Then } \exists j \in \nats \ldotp \Phi \step[j] \failed
        \end{array}
      \]
    \end{enumerate}
  \end{enumerate}
\end{definition}

  We require a global capability for invoking malloc (because if the capability were local, then a program with access to malloc would have to give up this access when invoking untrusted code, \ref{item:malloc-spec:eg}).
  The capability is assumed to have enter permission (\ref{item:malloc-spec:eg}), so malloc can protect its internal state even when the capability is shared with untrusted code.

  In addition to these syntactic requirements, we also specify standard behavioural expectations of malloc.
  Intuitively, we require that when malloc is invoked with a length argument, then it will return a capability for a fresh piece of memory of that size.
  It should be fresh in the sense that malloc has not already allocated any of that memory and will not do so in the future.
  Also, when invoked with a nonsensical length argument such as a negative integer or a capability, malloc should simply fail.
  However, formulating these requirements is harder than one might expect.
  The problem is that a realistic implementation of malloc needs to rely on internal state that changes after every invocation and relies on invariants on that state.
  We can only expect that malloc behaves according to its specification if its internal state satisfies the current state of the invariants in an executing system.
  We express this in terms of the semantic model from \sectionname~\ref{sec:logical-relation}.

To allow malloc implementations to rely on internal state and invariants for
that state, we assume an initial region for malloc (\ref{item:malloc-spec:reg}). The region is assumed to be
permanent (since safety of the global malloc capability will depend on its
presence, \ref{item:malloc-spec:perm}).
% DOMI: I don't understand what the following uncommented text means
% This also means that the malloc specification should guarantee that
% use of malloc is safe at any point of execution no matter who has invoked it. To
% this end, the requirement on the initial malloc region should also hold true in
% any private future world.
% priv future means that malloc should not depend on local capabilities?
Furthermore, we want to express that malloc does not depend on any other memory
than its own internal state. This is expressed by a restriction on the malloc
region's state interpretation function which (as explained in
\sectionname~\ref{subsec:worlds}) defines what memory segments it permits in a
given world. We require that the accepted memory segments only depend on the
presence of the malloc region itself, i.e.\ in any world the same memory
segments are accepted if we remove all regions except the malloc region. This
property should continue to hold throughout execution, so it must hold true for
any private future region of the initial malloc region (\ref{item:malloc-spec:independent}).

The malloc specification also dictates what malloc should do when invoked in a
memory with its internal state valid according to the malloc region (some future
evolution of the initial malloc region). If malloc is invoked with an invalid
length argument (that is, a negative integer or a capability), then we simply
require malloc to fail (\ref{item:malloc-spec:fail}). This part of the
specification does not actually rely on the malloc region: for simplicity, we
assume malloc does not need its internal state to check the argument. When
malloc is invoked with a valid length (\ref{item:malloc-spec:correct-exec}),
then it should return a fresh memory segment of the correct length. This segment
is required to come from the footprint of malloc, i.e.\ the memory owned by the
malloc region before the call. After malloc returns, we require the malloc
region to have evolved (according to the public future world relation) to a new
state where the new footprint is disjoint from the allocated memory. This
implies that future invocations of malloc can never return previously-allocated
memory.

@@

As described previously, the specification of malloc ensures that malloc has no
capabilities pointing \textit{out} of malloc. It does not, however, say anything
about capabilities that point \textit{in} to malloc. If we want to be able to
trust malloc, we obviously cannot allow arbitrary capabilities to point in to
it. We have chosen to keep the malloc specification simple and let this
assumption be in the lemmas that use malloc. It is sufficient for these lemmas
to require that there are no outside capabilities for malloc in the initial
configuration as capabilities cannot appear out of thin air, and the malloc
specification makes sure that capabilities are not leaked.

Malloc should not just be available to trusted programs but also to possibly
malicious programs. This is safe as it follows from the specification that the malloc
capability is always safe in a world with the malloc region:
\begin{lemma}[Malloc is safe to pass to adversary]
  \label{lem:malloc-in-vr}
  For all $c_\malloc$ that satisfies the specification for malloc with region $\iota_{\malloc,0}$,  if $W(r) \futurestr \iota_{\malloc,0}$, then
  $\npair{c_\malloc} \in \stdvr(W)$ for all $n$.
\end{lemma}
